---
title: "Computational Statistics - Lab 06"
author: "Maximilian Pfundstein"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    toc: true
    toc_float: false
    number_sections: false
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, include = TRUE, eval = TRUE)
library(ggplot2)
library(knitr)
library(gridExtra)
```

# Question 1: Genetic Algorithm

In this assignment, you will try to perform one-dimensional maximization with the help of a
genetic algorithm.

**Task:** Define the function

$$f(x) = \frac{x^2}{e^x}-2e^{\frac{-9sin(x)}{x^2+x+1}}$$
```{r}

f = function(x) {
  left = x^2/exp(x)
  exponent = (-9*sin(x))/(x^2+x+1)
  right = 2*exp(exponent)
  return(left-right)
}

```

**Task:** Define the function `crossover()`: for two scalars `x` and `y` it returns their "kid"" as `(x+y)/2`

```{r}

crossover = function(x, y) return((x+y)/2)

```

**Task:** Define the function `mutate()` that for a scalar `x` returns the result of the integer division $x^2 mod 30$. (Operation mod is denoted in R as `%%`)

```{r}

mutate = function(x) return((x^2)%%30)

```


**Task:** Write a function that depends on the parameters `maxiter` and `mutprob` and:

a. Plots function `f` in the range from 0 to 30. Do you see any maximum value?
b. Defines an initial population for the genetic algorithm as $X = (0, 5, 10, 15, . . . , 30)$
c. Computes vector `Values` that contains the function values for each population point.
d. Performs `maxiter` iterations where at each iteration
    1. Two indexes are randomly sampled from the current population, they are further used as parents (use `sample()`).
    2. One index with the smallest objective function is selected from the current population, the point is referred to as victim (use `order()`).
    3. Parents are used to produce a new kid by crossover. Mutate this kid with probability `mutprob` (use `crossover()`, `mutate()`).
    4. The victim is replaced by the kid in the population and the vector `Values` is updated.
    5. The current maximal value of the objective function is saved.
e. Add the final observations to the current plot in another colour.

**Answer:**

**`a:`** Let's have a look at the plot.

```{r, echo = FALSE}

sequence = seq(from = 0, to = 30, by = 1)
f.sequence. = f(sequence)
df = data.frame(sequence, f.sequence.)

ggplot(df) +
  geom_line(aes(x = sequence, y = f.sequence.), color = "#C70039") +
  labs(title = "f(x)", y = "f(x)", x = "x") +
  theme_minimal()

```

We see as maximum value at `x`:

```{r, echo = FALSE}

print(sequence[which.max(f.sequence.)])

```

With `f(x)`:

```{r}

print(max(f.sequence.))

```

**`b:`** Let's define the initial population:

```{r}

X = seq(from = 0, to = 30, by = 5)

```

**`c:`** Let us create the `Values`.

```{r}

Values = f(X)

```

**`d:`** Let's create a for loop that is performing `maxiter` iterations as maximum.

```{r}

maxiter = 100
mutprob = 0.05
best_individual = NaN

for (i in 1:maxiter) {
  
  # 1)
  parents = sample(1:length(X), size = 2)
  
  # 2) I don't know why we should order here!?
  victim = which.min(Values)
  
  # 3)
  child = crossover(X[parents][1], X[parents][2])
  if (mutprob > runif(n = 1, min = 0, max = 1)) {
    child= (mutate(child))
  }
  
  # 4)
  X[victim] = child
  Values[victim] = f(child)
  
  # 5)
  best_individual = list(cx = X[victim], cy = Values[victim])
}

```

**`e:`** Let's ad the final value to the plot.

```{r, echo = FALSE}

df = data.frame(sequence, f.sequence.)
best_individual = as.data.frame(best_individual)

ggplot(df) +
  geom_line(aes(x = sequence, y = f.sequence.), color = "#C70039") +
  geom_point(aes(x = cx, y = cy ), data = best_individual, color = "black",
            fill = "white", shape = 21, size = 3, stroke = 2) +
  labs(title = "f(x)", y = "f(x)", x = "x") +
  theme_minimal()

```


**Task:** Run your code with different combinations of `maxiter = 10, 100` and `mutprob= 0.1, 0.5, 0.9`. Observe the initial population and final population. Conclusions?

# Question 2: EM Algorithm

The data file `physical.csv` describes a behavior of two related physical processes $Y = Y (X)$
and $Z = Z(X)$.

**Task:** Make a time series plot describing dependence of Z and Y versus X. Does it seem that two processes are related to each other? What can you say about the variation of the response values with respect to X?

**Task:** Note that there are some missing values of Z in the data which implies problems in estimating models by maximum likelihood. Use the following model

$$Y_i \sim e^{x_i/\lambda}, \quad Z_i \sim e^{x_i/2\lambda}$$
where $\lambda$ is some unkown parameter.
**The goal is to derive an EM algorithm that estimates $\lambda$.**

**Task:** Implement this algorithm in `R`, use $λ_0 = 100$ and convergence criterion "stop if the change
in $\lambda$ is less than `0.001`". What is the optimal $\lambda$ and how many iterations were required to
compute it?

**Task:** Plot $E[Y]$ and $E[Z]$ versus X in the same plot as Y and Z versus X. Comment whether the computed $\lambda$ seems to be reasonable.


# Source Code

```{r, ref.label=knitr::all_labels(), echo = TRUE, eval = FALSE, results = 'show'}

```

